---
title: "Pathogen Exposed Sample Random Forest Classification"
author: "Timothy Smyth"
date: "2024-04-01"
output: html_document
---

# Pathogen Exposed Sample Random Forest Classification

### This file employs the random forest models generated in 'Random Forest' to classify bacterially or virally exposed MDM samples as M0, LPS, IFNy, or LPS+IFNy. This seeks to determine which polarization method best resembles MDMs following various pathogen exposures.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Environment setup

```{r message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

library(tibble)
library(tidyverse)
library(dplyr)
library(caTools)
library(caret)
library(ranger)
library(ROCR)
library(ConfusionTableR)
library(ggplot2)
library(RColorBrewer)
library(dendextend)
library(edgeR)
```

### Load and format count data for pathogen exposed samples

Input data needed for this process are the count and metadata of pathogen exposed samples, batch corrected and sample selected data from the original data set, and the three random forest models

```{r}
load("Infected.RData")
load("Infected_Metadata.RData")
load("Batched_Data_Select_Method.RData")

Infected_df %>% count(Agent)

Infected_df <- Infected_df %>% rownames_to_column() 

# Filter out samples with less than 5 samples and remove unuseable groups
Infected_df <- Infected_df %>%
  na.omit() %>%
  group_by(Agent) %>%
  filter(n() >= 5) %>% 
  data.frame()

Infected_df %>% count(Agent)

Infected_df$series_id <- gsub(",", "_", Infected_df$series_id) 

# Isolate sample info and set to factor
names <- Infected_df$rowname
info <- as.data.frame(Infected_df[, 2:4], row.names = names)
Infected_df <- Infected_df %>% subset(select = -c(1:4))
Infected_df <- as.data.frame(sapply(Infected_df, as.numeric))

# This changes gene name formatting for some reason
# This is important for filtering genes before limma-voom DEG calculation
Infected_df <- t(data.frame(Infected_df, row.names = names))

# Convert into a matrix
counts <- as.matrix(Infected_df)

Agent <- factor(info$Agent, 
                levels = c('Listeria',
                           'L. pneumophila',
                           'Mycobacterium smegmatis',
                           'Mycobacterium tuberculosis',
                           'Salmonella typhimurium',
                           'Staphylococcus epidermidis',
                           'Yersinia pseudotuberculosis',
                           
                           'Chikungunya Virus',
                           'Ebola Virus',
                           'HIV',
                           'IAV',
                           'Reston Virus',
                           'Zika Virus'))
```

### Remove genes that were removed from the original data set and convert count data to log count per million data

```{r}
# Create DGEList object
d0 <- DGEList(Infected_df, 
              group = Agent)

# Calculate normalization factor
d0 <- calcNormFactors(d0)

# Remove filtered genes
d0 <- d0[colnames(Batched_MDM_df[5:ncol(Batched_MDM_df)]), ]

# Calculate normalization factor
d0 <- calcNormFactors(d0)

# Calculate logCPM of filtered gene set
lcpm <- cpm(d0, log = TRUE)

# Add polarization information and series_ID
lcpm <- as.data.frame(t(lcpm))

lcpm <- data.frame(info, lcpm)
```

### Select samples using dendrogram clustering

```{r}
list <- as.list(unique(as.character(lcpm$Agent)))

dend <- lapply(list, function(x){
  
  tmp <- lcpm %>% subset(Agent == x)
  names <- rownames(tmp)
  tmp_series <- tmp$series_id
  
  # Remove metadata, convert count data to numeric, and set sample to row names
  tmp <- tmp %>% 
    subset(select = -c(1:3)) %>% 
    as.data.frame(row.names = names)
 
  # Calculate sample correlation and create dendrograms
  tmp_dend <- cor(t(tmp))
  tmp_dend <- as.dendrogram(hclust(as.dist(1 - tmp_dend)))
  
  # Isolate each unique GEO series ID
  useries = unique(tmp_series)
  
  # Match unique series to series list and record row location of each series
  series_match = useries[match(tmp_series, useries)]
  
  # Set colors
  colos <- colorspace::rainbow_hcl(length(tmp_series), 
                                   c = 160, 
                                   l  = 50)
  # Set colors to series ID
  names(colos) = tmp_series
  
  # Set matched colors
  series_color <- colos[series_match]
  
  # Create clusters
  clu = cutree(tmp_dend, 
               h = 0.25) # Height for cut (1 - correlation coefficient)
  
  # Set colors of labels
  labels_colors(tmp_dend) <- series_color[order.dendrogram(tmp_dend)]
  
  # Create dendrograms
  tmp_dend <- color_branches(tmp_dend, 
                             h = 0.25) # Height of cut (1 - correlation coefficient)
  
  # Set plot dimmensions
  par(mar = c(4,1,1,12))
  
  # Plot dendrograms
  plot(tmp_dend,
       main = as.character(x),
       horiz = TRUE) +
    
    abline(v = 0.25, lty = 2)
  
  # Create color bar to show cluster and series ID of samples
  colored_bars(cbind(clu, 
                     series_color), 
               tmp_dend, 
               rowLabels = c("Cluster", "Series"), 
               horiz = TRUE)
  
  # Isolate samples from single polarization method
  tmp <- lcpm %>% subset(Agent == x)
  
  # Add cluster number information to data
  tmp <- cbind(clu, tmp)
  colnames(tmp)[1] <- 'cluster'
  tmp <- tmp %>% select(ID, everything())
  
  # Determine the largest cluster and isolate samples in that cluster
  largest_cluster = names(rev(sort(table(tmp$cluster))))[1]
  samples = which(tmp$cluster == largest_cluster)
  tmp[samples, ]
  
})

Selected_Infected_df <- do.call(rbind, dend) %>% subset(select = -c(2))
info <- info[rownames(Selected_Infected_df), ]

info %>% count(Agent)
```

### Remove batch effect with removeBatchEffect

```{r}
# Create model matrix to remove batch effects
Design <- model.matrix(~ Agent, 
                       data = info)

# Set info to factor for removeBatchEffect
info$series_id <- factor(info$series_id)

Batched_lcpm <- t(Selected_Infected_df[4:ncol(Selected_Infected_df)])

# Requires samples as columns
Batched_lcpm <- limma::removeBatchEffect(Batched_lcpm,
                                         batch = info$series_id,
                                         design = Design)

Batched_lcpm <- t(Batched_lcpm)

save(file = 'Batched_Infected.RData', 
     Batched_lcpm, 
     info, 
     Infected)
```

### Perform random forest classification using the random forest models

```{r}
load("RF_Model.RData") # Original model with all significant genes
load("Top_10000_Pruned_RF_Model.RData") # Model with top 10000 genes by permutation importance of original model
load("Top_1000_Pruned_RF_Model.RData") # Model with top 1000 genes by permutation importance of top 10000 model

list <- list(classifier_RF, 
             Top_10000_classifier_RF_pruned,
             Top_1000_classifier_RF_pruned)

lapply(list, function(x){
  
  # Predict
  p1 <- predict(x, Batched_lcpm)
  
  # Add sample information to prediction outcomes
  prediction <- cbind(rownames(Batched_lcpm), p1[["predictions"]])
  colnames(prediction) <- c('sample', 'Prediction')
  prediction <- data.frame(prediction)
  
  info <- info %>% rownames_to_column()
  colnames(info)[1] <- 'sample'
  
  prediction <- merge(info, prediction, by = 'sample')
  
  # Change prediction number to ID matching prediction
  # This corresponds to the factor levels of the RF models
  prediction$Prediction[prediction$Prediction == 1] <- 'M0'
  prediction$Prediction[prediction$Prediction == 2] <- 'IFN'
  prediction$Prediction[prediction$Prediction == 3] <- 'LPS'
  prediction$Prediction[prediction$Prediction == 4] <- 'IFN_LPS'
  
  prediction$Prediction <- factor(prediction$Prediction, 
                                  levels = c('M0', 
                                             'IFN', 
                                             'LPS',
                                             'IFN_LPS'))
  
  prediction$Agent <- factor(prediction$Agent, 
                             levels = c('Listeria',
                                        'L. pneumophila',
                                        'Mycobacterium smegmatis',
                                        'Mycobacterium tuberculosis',
                                        'Salmonella typhimurium',
                                        'Staphylococcus epidermidis',
                                        'Yersinia pseudotuberculosis',
                                        
                                        'Chikungunya Virus',
                                        'Ebola Virus',
                                        'HIV',
                                        'IAV',
                                        'Reston Virus',
                                        'Zika Virus'))
  
  # Plot the results
  ggplot(data = prediction, 
       aes(x = Agent, 
           y = Prediction)) +
  
  geom_jitter(aes(color = Prediction), 
              width = 0.2) +
  
  theme(axis.text.x = element_text(size = 12, 
                                   angle = 45, 
                                   hjust = 1), 
        
        axis.text.y = element_text(size = 12), 
        
        axis.title = element_text(size = 16))
})
```

