---
title: "GSVA Analysis"
author: "Timothy Smyth"
date: "2024-03-27"
output: html_document
---

# Gene Ontology (GO) Biological Processes (BP) Term Enrichment and Gene Set Variation Analysis (GSVA)

### This file performs GO BP term enrichment and subsequent Gene Set Variation Analysis (GSVA) of enriched terms. GSVA allows gene sets, which contain genes associated with specific biological processes, to be tested for overall expression and collapsed into a single enrichment value. Following GSVA calculations, differential expression of these single GSVA values can be performed using limma-voom. Top enriched gene sets can then be extracted and compared between groups using a heatmap describing GSVA values and a yes/no significance heatmap.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Prepare environment

```{r message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

library(tibble)
library(tidyverse)
library(dplyr) 
library(topGO)
library(org.Hs.eg.db)
library(GO.db)
library(biomaRt)
library(GSVA)
library(GSEABase)
library(edgeR)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
```

### Prepare GO IDs using the human ENSEMBL gene BioMart database

```{r}
# load gene universe for GO term assignment
load("Gene_universe.RData")

# Select human ENSEMBL gene BioMart database and data set
db = useMart('ENSEMBL_MART_ENSEMBL',
             dataset = 'hsapiens_gene_ensembl', 
             host = "https://www.ensembl.org"
             )

# Assign GO IDs to genes
go_ids = getBM(attributes = c('go_id', 
                              'name_1006',
                              'ensembl_gene_id', 
                              'namespace_1003'), 
               filters = 'ensembl_gene_id', 
               values = gene_universe, 
               mart = db)
```

### Perform GO BP term enrichment

```{r}
# Load count data
load("Top_1000_RF_Genes.RData")

MDM_results <- t(MDM_df[, top_1000])

# Create annotation list for TopGo object
gene_2_GO = unstack(go_ids[, c(1, 3)])

# Turn off scientific notation
options(scipen=999)

# Save gene names
names <- rownames(MDM_results)

# remove any candidate genes without GO annotation
keep = names %in% go_ids[, 3]
keep = which(keep == TRUE)
names = names[keep]

# make named factor showing which genes are of interest
geneList = factor(as.integer(gene_universe %in% names))
names(geneList) = gene_universe

# Create TopGo object
GOdata = new('topGOdata', 
             ontology = 'BP', 
             allGenes = geneList, 
             annot = annFUN.gene2GO, 
             gene2GO = gene_2_GO)

# Define test using the weight01 algorithm with fisher
weight_fisher_result = runTest(GOdata, 
                               algorithm = 'weight01', 
                               statistic = 'fisher') 

# Generate a table of results
allGO = usedGO(GOdata)

all_res = GenTable(GOdata, 
                   weightFisher = weight_fisher_result, 
                   orderBy = 'weightFisher', 
                   topNodes = length(allGO), 
                   numChar = 1000)

# Performing BH correction on our p values
p.adj = round(p.adjust(all_res$weightFisher, 
                       method = "BH"), digits = 4)

# Add adjusted p value to results and order based on adjusted p value
all_res_final = cbind(all_res, p.adj)
all_res_final = all_res_final[order(all_res_final$p.adj),]

###############################################################################
###############################################################################

# Annotated:
# number of genes (in our gene list) that are annotated with the term

# Significant:
# Number of significantly DE genes annotated with that term (i.e. genes where geneList = 1)

# Expected:
# Under random chance, number of genes that would be expected to be significantly DE and annotated with that term

# p.value:
# P-value from Fishers Exact Test, testing for association between significance and pathway membership.

###############################################################################
###############################################################################
```

### Assign genes affiliated to each GO BP term to the linked term

```{r}
# Isolate significant GO pathways and add matching gene names for GSVA below
Sig_GO <- all_res_final %>% subset(weightFisher < 0.05)
Sig_GO$Genes <- ''

tmp <- as.list(Sig_GO$GO.ID)

tmp2 <- lapply(tmp, function(y){
  
  tmp3 <- go_ids %>% subset(go_id == y)
  tmp3 <- as.character(tmp3$ensembl_gene_id)
  
})

Sig_GO$Genes <- tmp2
```

### Create gene sets and perform GSVA

```{r}
save <- MDM_df[1:3]

# Store treatment group data
phenoData <- new("AnnotatedDataFrame",
                 data = save)

# Create expression set containing count data and metadata
lcpm <- ExpressionSet(assayData = as.matrix(t(MDM_df[4:ncol(MDM_df)])),
                      phenoData = phenoData)

Gene_sets <- list()

for(i in 1:length(Sig_GO[["Term"]])){
  
  Gene_sets[[i]] <- GeneSet(Sig_GO[["Genes"]][[i]], 
                            setName = Sig_GO[["Term"]][[i]])
  
}

names(Gene_sets) <- Sig_GO[["Term"]]

# Run GSVA on each pathways
GSVA_results <- gsva(lcpm,
                     GeneSetCollection(Gene_sets),
                     verbose = FALSE)

GSVA_results <- cbind(GSVA_results@phenoData@data,
                      t(GSVA_results@assayData[["exprs"]]))
```

### Perform differential expression analysis of GSVA scores

```{r}
# Isolate metadata
info <- GSVA_results[, 1:3]

ID <- factor(info$ID,
             level = c ("M0", 
                        "LPS_IFN", "LPS", "IFN"))

# Convert into a matrix
counts <- as.matrix(t(GSVA_results[4:ncol(GSVA_results)]))

# Create a model with ID as factor
model <- model.matrix(~0 + ID)

colnames(model)[1:4] <- c("M0", 
                          "LPS_IFN", "LPS", "IFN")

# Remove any columns which have no values (i.e. unrepresented in study design)
model <- model[, colSums(model != 0) > 0]

# Check if any coefficients in model cannot be estimated
# Return value of "NULL" means model has full column rank
nonEstimable(model)

# Fit linear model using weighted least squares for genes
fit <- lmFit(counts, 
             model)
```

### Calculate pathway differential expression p-values

```{r}
# Make a list of contrasts
list <- list(LPS_IFN_vs_M0 = makeContrasts(LPS_IFN-M0, 
                                           levels = model),
             
             LPS_vs_M0 = makeContrasts(LPS-M0, 
                                       levels = model),
             
             IFN_vs_M0 = makeContrasts(IFN-M0,
                                       levels = model), 
             
             ################
             
             LPS_IFN_vs_LPS = makeContrasts(LPS_IFN-LPS, 
                                           levels = model),
             
             LPS_IFN_vs_IFN = makeContrasts(LPS_IFN-IFN, 
                                           levels = model),
             
             LPS_vs_IFN = makeContrasts(LPS-IFN, 
                                           levels = model))

############

# Calculate DEGs
DEGs <- lapply(list, function(x){
  
  tmp <- contrasts.fit(fit, x)
  tmp <- eBayes(tmp)
  
  topTable(tmp, sort.by = "P", n = Inf)
  
})

# Results are the following information:

# logFC: log0 fold change of group1/group0
# AveExpr: Average expression across all samples, in log0 CPM
# t: logFC divided by its standard error
# P.Value: Raw p-value (based on t) from test that logFC differs from 0
# adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value
# B: log-odds that gene is DE

DEGs <- lapply(DEGs, function(x){
  
  tmp <- data.frame(cbind(x[["logFC"]], 
                          x[["P.Value"]],
                          x[["adj.P.Val"]],
                          x["AveExpr"]),
                    row.names = rownames(x)) 
  
  colnames(tmp) <- c('log2FC', 'P.Value', 'adj.P.Val', 'Expression')
  tmp
  
})
  
list <- as.list(names(DEGs)[1:3])
```

### Isolate the top 20 pathway names for each comparison versus M0 and calculate mean GSVA scores by group

```{r}
# Isolate the pathway names which are significantly differentially expressed
pathway_names <- lapply(list, function(x){
  
  tmp <- DEGs[[x]] %>% arrange(adj.P.Val) %>% subset(adj.P.Val < 0.05)
  tmp <- rownames(tmp)
  
  # If there are any pathways differentially expressed, isolate the top 20
  # Otherwise, mark the comparison with 'Remove'
  if(length(tmp) > 0){
    tmp <- tmp[1:20]
  }
  else{
    tmp <- 'Remove'
    }
  
})

# Remove comparisons without DE pathways
pathway_names <- pathway_names[pathway_names != 'Remove']
pathway_names <- as.list(unique(unlist(pathway_names)))
pathway_names <- pathway_names[pathway_names != 'NA'] # Remove NA results
pathway_names <- as.character(unique(unlist(pathway_names)))

# Isolate GSVA results for DE pathways
DEG_results <- GSVA_results[, pathway_names]
info <- info[rownames(DEG_results), ]  
DEG_results <- data.frame(info, DEG_results)

# Calculate mean GSVA scores
DEG_results <- DEG_results %>% 
  group_by(ID) %>% 
  summarize(across(everything(), mean))
  
DEG_results <- data.frame(DEG_results[4:ncol(DEG_results)], row.names = DEG_results$ID)
colnames(DEG_results) <- pathway_names
colnames(DEG_results) <- gsub('\\.', ' ', colnames(DEG_results)) %>% stringr::str_to_title() # Reformat pathway names
```

### Generate a heatmap of mean GSVA values for top pathways

```{r}
# Set color palette
myCol <- colorRamp2(c(-1, 0, 1), hcl_palette = "Vik")

DEG_results <- DEG_results[c('M0', 'LPS', 'IFN', 'LPS_IFN'), ]

hmap <- Heatmap(t(DEG_results),
                
                # split the columns / rows
                row_split = colnames(DEG_results),
                row_gap = unit(0, 'mm'),
                rect_gp = gpar(col = "black"),
                border = FALSE,                
                width = nrow(DEG_results) * unit(50, 'mm'),
                height = ncol(DEG_results) * unit(15, 'mm'),
                
                name = 'GSVA\nEnrichment Score',
                
                col = myCol,
                
                # Add mean GSVA values
                # Change text color based on value
                cell_fun = function(j, i, x, y, width, height, fill) {
                  grid.text(sprintf("%.2f", t(DEG_results)[i, j]), x, y, gp = gpar(fontsize = 20, 
                                                                           col = if_else(t(DEG_results)[i, j] < -0.75, 
                                                                                         'white', 
                                                                                         'black'),
                                                                           fontface = 'bold'))},
                
                # parameters for the color-bar that represents gradient of expression
                heatmap_legend_param = list(
                  color_bar = 'continuous',
                  at = c(-1, -0.5, 0, .5, 1),
                  legend_direction = 'horizontal',
                  legend_width = unit(12, 'cm'),
                  legend_height = unit(12, 'cm'),
                  title_position = 'topcenter',
                  title_gp = gpar(fontsize = 24, fontface = 'bold'),
                  labels_gp = gpar(fontsize = 20, fontface = 'bold')),
                
                # row (gene) parameters
                cluster_rows = FALSE,
                show_row_dend = FALSE,
                row_title = NULL,
                row_title_side = 'left',
                row_title_gp = gpar(fontsize = 12,  fontface = 'bold'),
                show_row_names = TRUE,
                row_names_gp = gpar(fontsize = 24, fontface = 'bold'),
                row_names_side = 'left',
                row_dend_width = unit(25,'mm'),
                
                # column (sample) parameters
                cluster_columns = TRUE,
                show_column_dend = TRUE,
                show_column_names = TRUE,
                column_title = NULL,
                column_names_gp = gpar(fontsize = 30, fontface = 'bold'),
                column_names_rot = 45,
                column_names_max_height = unit(20, 'cm'),
                column_dend_height = unit(25,'mm'))

png(file = "GSVA_DE_Heatmap.png", height = 3000, width = 2500)

draw(hmap,
     heatmap_legend_side = 'bottom', 
     row_dend_side = 'right')

dev.off()
```

### Generate a heatmap describing whether pathway GSVA scores are significantly different between M1 group

```{r}
list <- as.list(names(DEGs)[4:6])

# Mark pathways as significant or non-significant DE
significant <- lapply(list, function(x){
  
  tmp <- DEGs[[x]][pathway_names, ]
  tmp$Sig <- if_else(tmp$adj.P.Val >= 0.05, 'Not_Significant', 'Significant')
  tmp$Sig

})

significant <- do.call(cbind, significant)
colnames(significant) <- list
rownames(significant) <- pathway_names
rownames(significant) <- gsub('\\.', ' ', rownames(significant)) %>% stringr::str_to_title()
significant <- significant[colnames(DEG_results), ]

hmap <- Heatmap(significant,
                
                # split the columns / rows
                row_split = rownames(significant),
                
                column_split = factor(colnames(significant), 
                                      levels = c("LPS_IFN_vs_LPS", 
                                                 "LPS_IFN_vs_IFN",
                                                 "LPS_vs_IFN" ,
                                                 "IFN_vs_LPS")),
                
                row_gap = unit(0, 'mm'),
                column_gap = unit(0, 'mm'),
                rect_gp = gpar(col = "black"),
                border = FALSE,
                width = nrow(DEG_results) * unit(50, 'mm'),
                height = ncol(DEG_results) * unit(15, 'mm'),
                
                col = c('tomato', 'lightblue'),
                
                name = ' ',
                
                # row (gene) parameters
                cluster_rows = FALSE,
                show_row_dend = FALSE,
                row_title = NULL,
                row_title_side = 'left',
                row_title_gp = gpar(fontsize = 12,  fontface = 'bold'),
                show_row_names = TRUE,
                row_names_gp = gpar(fontsize = 24, fontface = 'bold'),
                row_names_side = 'right',
                row_dend_width = unit(25,'mm'),
                
                # column (sample) parameters
                cluster_columns = FALSE,
                cluster_column_slices = FALSE,
                show_column_dend = FALSE,
                show_column_names = TRUE,
                column_title = NULL,
                column_names_gp = gpar(fontsize = 30, fontface = 'bold'),
                column_names_rot = 45,
                column_names_max_height = unit(20, 'cm'),
                column_dend_height = unit(25,'mm'))

png(file = "GSVA_DE_Significance_Heatmap.png", height = 3000, width = 2500)

draw(hmap,
     heatmap_legend_side = 'bottom')

dev.off()
```

