---
title: "Gene and Sample Filtering"
author: "Timothy Smyth"
date: "2024-03-27"
output: html_document
---

# Gene and Sample filtering of Monocyte-Derived Macrophage Samples

### This file takes the count data from all M0, M1, and M2 samples isolated from "MDM Selectiona and Sorting" and filters genes and samples for differential expression and random forest modeling. Genes are filtered with the edgeR function filterByExpr() using default parameters which determines whether each gene has sufficiently large counts across a sufficient number of samples. Samples are filtered using dendrogram clustering, retaining samples with a pearson correlation of at least 0.75 within each polarization method. Resulting data is either saved as raw count data for differential expression or corrected for batch effect for use in random forest modeling and further downstream analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Environment setup

```{r message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

# Load packages
library(tidyverse) # for data cleaning
library(dplyr) 
library(broom)
library(tibble)
library(edgeR)
library(RColorBrewer) # for color palette
library(dendextend)
library(factoextra)
library(limma)
```

### Load and format count data

```{r}
# Load MDM data
load("MDM_Data_by_Polarization.RData")

# Isolate polarization methods (termed ID in data frame) of interest
# These methods were selected due to their widespread use to induce M1 or M2 polarization states in the literature
MDM_df <- MDM_df %>% subset(ID == 'M0' |
                              ID == 'LPS' |
                              ID == 'IFN' |
                              ID == 'LPS_IFN' |
                              ID == 'IL4' |
                              ID == 'IL10' | 
                              ID == 'IL13' |
                              ID == 'IL4_IL13')

head(MDM_df)

# Move 'polarization' (referring to M0, M1, or M2 polarization state) to the first column
MDM_df <- MDM_df %>% select(polarization, everything())

# Set series_id and polarization state to a factor
MDM_df$series_id <- gsub(",", "_", MDM_df$series_id) %>% factor()
MDM_df$polarization <- factor(MDM_df$polarization)
MDM_df$ID <- factor(MDM_df$ID)

# Isolate sample info and set to factor
info <- MDM_df %>% subset(select = c(1:3))

# Save sample name
names <- rownames(MDM_df)

# Subset data to remove metadata, set count data to numeric, and set to a data frame
MDM_df <- MDM_df %>% subset(select = -c(1:3))
MDM_df <- sapply(MDM_df, as.numeric)
MDM_df <- t(data.frame(MDM_df, row.names = names))

# Define GEO series as a factor
series <- factor(info$series_id)
polarization <- factor(info$polarization, levels = c('M0', 'M1', 'M2'))
pol_method <- factor(info$ID, 
                     levels = c('M0', 
                                'LPS', 'IFN', 'LPS_IFN',  
                                'IL4', 'IL10', 'IL13', 'IL4_IL13'))
```

### Filter genes without sufficiently large counts across samples using filterByExpr

```{r}
# Create DGEList object
d0 <- DGEList(as.matrix(MDM_df), 
              group = pol_method)

# Calculate normalization factor
d0 <- calcNormFactors(d0)

# Use EdgeR function to automatically select genes to retain
# This gives logical values to keep or remove
keep.exprs <- filterByExpr(d0, 
                           group = pol_method)

# Remove filtered genes
d0 <- d0[keep.exprs, ]

# Recalculate normalization factor
d0 <- calcNormFactors(d0)

# Calculate logCPM of filtered gene set
lcpm <- cpm(d0, log = TRUE)

# Isolate names of genes retained and save for future analysis
gene_universe <- rownames(lcpm)

save(file = 'Gene_universe.RData', 
     gene_universe)

# Add polarization information and series_ID
lcpm <- as.data.frame(t(lcpm))
lcpm <- cbind(info, lcpm)

colnames(lcpm)[1:3] <- c('polarization', 
                         'series_id', 
                         'ID')
```

### Fitler samples through dendrogram clustering. Samples are split into groups based on the identified polarization method and pearson correlations are calculated. Dendrograms are constructed based on these between sample correlation coefficients and dendrograms are cut at heights of 0.25, corresponding to a pearson correlation of at least 0.75. Samples from the largest dendrogram cluster are retained for downstream analysis

```{r}
# Isolate polarization method names for clustering
list <- as.list(unique(as.character(lcpm$ID)))

dend <- lapply(list, function(x){
  
  # Isolate samples from single polarization method
  tmp <- lcpm %>% subset(ID == x) 
  names <- rownames(tmp)
  tmp_series <- tmp$series_id
  
  # Remove metadata, convert count data to numeric, and set sample to row names
  tmp <- tmp %>% 
    subset(select = -c(1:3)) %>% 
    as.data.frame(row.names = names)
 
  # Calculate sample correlation and create dendrograms
  tmp_dend <- cor(t(tmp))
  tmp_dend <- as.dendrogram(hclust(as.dist(1 - tmp_dend)))
  
  # Isolate each unique GEO series ID
  useries = unique(tmp_series)
  
  # Match unique series to series list and record row location of each series
  series_match = useries[match(tmp_series, useries)]
  
  # Set colors
  colos <- colorspace::rainbow_hcl(length(tmp_series), 
                                   c = 160, 
                                   l  = 50)
  # Set colors to series ID
  names(colos) = tmp_series
  
  # Set matched colors
  series_color <- colos[series_match]
  
  # Create clusters
  clu = cutree(tmp_dend, 
               h = 0.25) # Height for cut (1 - correlation coefficient)
  
  # Set colors of labels
  labels_colors(tmp_dend) <- series_color[order.dendrogram(tmp_dend)]
  
  # Create dendrograms
  tmp_dend <- color_branches(tmp_dend, 
                             h = 0.25) # Height of cut (1 - correlation coefficient)
  
  # Set plot dimmensions
  par(mar = c(4,1,1,12))
  
  # Plot dendrograms
  plot(tmp_dend,
       main = as.character(x),
       horiz = TRUE) +
    
    abline(v = 0.25, lty = 2)
  
  # Create color bar to show cluster and series ID of samples
  colored_bars(cbind(clu, 
                     series_color), 
               tmp_dend, 
               rowLabels = c("Cluster", "Series"), 
               horiz = TRUE)
  
  # Isolate samples from single polarization method
  tmp <- lcpm %>% subset(ID == x)
  
  # Add cluster number information to data
  tmp <- cbind(clu, tmp)
  colnames(tmp)[1] <- 'cluster'
  tmp <- tmp %>% select(ID, everything())
  
  # Determine the largest cluster and isolate samples in that cluster
  largest_cluster = names(rev(sort(table(tmp$cluster))))[1]
  samples = which(tmp$cluster == largest_cluster)
  tmp[samples, ]
  
})

Selected_MDM_df <- do.call(rbind, dend) %>% 
  subset(select = -c(2)) # Remove cluster number

info <- info[rownames(Selected_MDM_df), ] # Remove info from removed samples
MDM_df <- t(MDM_df[keep.exprs, rownames(info)]) # Isolate raw count data for selected genes and samples

save(file = 'Selected_Count_Data.RData', 
     MDM_df,
     info)
```

### Perform batch correction using limma function removeBatchEffect using experimental Series_ID as the batch designation.

```{r}
info <- Selected_MDM_df[1:3]

# Create model matrix to remove batch effects
Design <- model.matrix(~ ID, 
                       data = info)

# Set series to factor for removeBatchEffect
info$series <- factor(info$series)

Batched_MDM_df <- t(Selected_MDM_df[4:ncol(Selected_MDM_df)])

# Remove batch effects caused by experimental series. 
# Requires samples as columns
Batched_MDM_df <- removeBatchEffect(Batched_MDM_df,
                                    batch = info$series,
                                    design = Design)

Batched_MDM_df <- data.frame(info, t(Batched_MDM_df))

save(file = 'Batched_Data_Select_Method.RData', 
     Batched_MDM_df,
     info)
```

### Create density plots for Supplemental Figure S1

```{r}
# Create density plots for sample data correlation at baseline, 
# following sample selection, and following batch correction
lapply(list, function(x){
  
  # Isolate pre-selected samples from single polarization method
  pre <- lcpm %>% subset(ID == x)
  pre <- pre[, 4:ncol(pre)]
  
  # Isolate post-selected samples from single polarization method
  post <- Selected_MDM_df %>% subset(ID == x)
  post <- post[, 4:ncol(post)]
  
  # Isolate batch corrected samples from single polarization method
  batch <- Batched_MDM_df %>% subset(ID == x)
  batch <- batch[, 5:ncol(batch)]

  # Plot density plots
  plot(density(cor(t(pre))), 
       lwd = 3, 
       main = paste0("Correlation of All ", x, " Samples"), 
       col = "red",
       ylim = c(0,30))
  
  lines(density(cor(t(post))), 
        lwd = 3, 
        main = paste0("Correlation of Selected ", x, " Samples"), 
        col = "blue")
  
  lines(density(cor(t(batch))), 
        lwd = 3, 
        main = paste0("Correlation of Batched ", x, " Samples"),
        col = "forestgreen")
  
  abline(v = 0.75, col = "black")
  
  legend("topleft", 
         legend = c("All Samples", "Selected Samples", "Batch Corrected Selected Samples"), 
         lty = 1, 
         lwd = 3, 
         col = c("red", "blue", "forestgreen"))
  
})
```

