---
title: "IPA Canonical Pathways"
author: "Timothy Smyth"
date: "2024-03-27"
output: html_document
---

# Ingenuity Pathway Analysis: Canonical Pathways

### This file describes further analysis of canonical pathway results from Ingenuity Pathway Analysis. Differentially expressed genes from 'Differential Expression by Polarization State' or 'Differential Expression by Polarization Method' were uploaded to Ingenuity Pathway Analysis with DEG cutoffs of BH adjusted p-values of 0.05 and log2 Fold Change values of >= 2 or <= -2. IPA core analysis was perfromed on each differential expression comparison (i.e. M1 vs M0) and the resulting analysis was downloaded as an excel file. Input for this file are the resulting IPA excel results.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Environment setup

```{r, message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

# Load packages
library(tidyverse) # for data cleaning
library(dplyr) 
library(readxl)
library(ComplexHeatmap)
library(circlize)
library(openxlsx)
library(eulerr)
```

### Load data

This requires additional setup to run. The working directory has to be set to the specific folder containing IPA results, and the working directory must also be added in paste0(). As a note, / is required at the end of the first section of paste0().

```{r}
# Create a list of folders containing IPA data
# All IPA folders should exist as separate sub folders in the same folder
Groupings <- c('By Polarization State', 
               'M1 vs M0', 
               'M1 vs M1', 
               'M2 vs M0', 
               'M2 vs M2')

# Create a list of the files from your target directory
file_list <- lapply(Groupings, function(DE_Group){
  
  # Define file path for data folders
    file_names <- list.files(path = paste0("---", 
                             DE_Group))
    
    Group_storage_list <- list()
    
    # Read each file into a separate data frame
    for(i in 1:length(file_names)){
      
      # Create a temporary data frame containing data from each file
      Tmp_df <- as.data.frame(read_excel(paste0('---',
                                                                 DE_Group, 
                                                                 '/',
                                                                 file_names[[i]]), 
                                                          col_types = 'text'))
      
      # Remove first 21 rows
      Tmp_df <- Tmp_df[-c(1:21), ]
      
      # Set rownames to sequential order of 1 to nrow
      row.names(Tmp_df) = seq(1, 
                              nrow(Tmp_df), 
                              by = 1) 
      
      # Save the temporary data frame to the storage list
      Group_storage_list[[i]] <- Tmp_df
      
      }
    
    # Set the names of the data frames to match the source file
    names(Group_storage_list) <- substr(file_names,
                                        1,
                                        nchar(file_names) - 4)
    
    # Save the data frames from each grouping to the same list
    Group_storage_list
  
})

# Rename the list to match the groups
names(file_list) <- Groupings
```

### Format IPA results for isolation of canonical pathway results

```{r}
IPA_Results <- lapply(Groupings, function(IPA_Grouping){
  
  IPA_Results <- lapply(file_list[[IPA_Grouping]], function(Individual_file){
    
    # ID the row and column number with the indicated text
    # These correspond to the IPA readouts
    Analysis <- which(Individual_file == 'Analysis', arr.ind = TRUE)
    
    # Subset from row 1 to 2 rows before 'Analysis' text (last data point)
    Ingenuity_Canonical_Pathways <- Individual_file[c(1:as.numeric(Analysis[1]-2)), 
                                                    c(1:ncol(Individual_file))] 
    
    # Set colnames to names in first row
    names(Ingenuity_Canonical_Pathways) <- lapply(Ingenuity_Canonical_Pathways[1, ], 
                                                  as.character) 
    
    # Remove first row which only contains colnames
    Ingenuity_Canonical_Pathways <- as.data.frame(Ingenuity_Canonical_Pathways[-1, ]) 
    
    # ID columns with blank names
    keep.cols <- names(Ingenuity_Canonical_Pathways) %in% c("") 
    
    # Remove columns without names
    Ingenuity_Canonical_Pathways <- Ingenuity_Canonical_Pathways[!keep.cols] 
    
    # Reset row names
    row.names(Ingenuity_Canonical_Pathways) = seq(1, 
                                                  nrow(Ingenuity_Canonical_Pathways), 
                                                  by = 1) 
    
    Ingenuity_Canonical_Pathways
    
    })
  
})

# Rename the list to match the groups
names(IPA_Results) <- Groupings
```

### Isolate canonical pathway results from formatted IPA data

```{r}
# Apply function to round z scores to 3 digits, reorder columns, and 
# subset by significantly differences (-log(p) > 1.3 = p<0.05)
Canonical <- lapply(Groupings, function(IPA_Grouping){
  
  Canonical <- lapply(names(IPA_Results[[IPA_Grouping]]), function(Individual_file){
  
    # Isolate the first position of the IPA results (Canonical Pathways)
    Ingenuity_Canonical_Pathways <- IPA_Results[[IPA_Grouping]][[Individual_file]]
    
    colnames(Ingenuity_Canonical_Pathways) <- c('Pathway', 
                                                'Neg.log.10.p', 
                                                'zScore', 
                                                'Ratio')
  
    # Set z score to numeric
    Ingenuity_Canonical_Pathways$zScore <- as.numeric(Ingenuity_Canonical_Pathways$zScore)
    
    # Round to 3 digits
    Ingenuity_Canonical_Pathways$zScore <- round(Ingenuity_Canonical_Pathways$zScore, 
                                                 3)
    
    # Reorganize data and keep only significantly different pathways
    Ingenuity_Canonical_Pathways <- Ingenuity_Canonical_Pathways[, c(1, 3, 4, 2)] %>% 
      subset(Ingenuity_Canonical_Pathways$Neg.log.10.p >= 1.3)
    
    # Set names of columns
    colnames(Ingenuity_Canonical_Pathways) <- c('Pathway',
                                                paste0(Individual_file, ' zScore'),
                                                paste0(Individual_file, ' Ratio'),
                                                paste0(Individual_file, ' Neg.log.10.p'))
    
    # Save the results to Canonical list
    Ingenuity_Canonical_Pathways
    
  })
  
})

# Rename the list to match the groups
names(Canonical) <- Groupings
```

### Isolate the top 10 canonical pathways by p-value of each comparison

```{r}
Top_pathways <- lapply(Groupings, function(IPA_Grouping){
  
  Top_pathways <- lapply(names(IPA_Results[[IPA_Grouping]]), function(Individual_file){
    
    # Isolate the first position of the IPA results (Canonical Pathways)
    Ingenuity_Canonical_Pathways <- IPA_Results[[IPA_Grouping]][[Individual_file]]
  
    colnames(Ingenuity_Canonical_Pathways) <- c('Pathway', 
                                                'Neg.log.10.p', 
                                                'zScore', 
                                                'Ratio')
    
    # Set to numeric
    Ingenuity_Canonical_Pathways$zScore <- as.numeric(Ingenuity_Canonical_Pathways$zScore)
    Ingenuity_Canonical_Pathways$Neg.log.10.p <- as.numeric(Ingenuity_Canonical_Pathways$Neg.log.10.p)
    Ingenuity_Canonical_Pathways$Ratio <- as.numeric(Ingenuity_Canonical_Pathways$Ratio)
    
    # Round to 3 digits
    Ingenuity_Canonical_Pathways$zScore <- round(Ingenuity_Canonical_Pathways$zScore, 3)
    
    # Reorganize data and keep only significantly different pathways
    Ingenuity_Canonical_Pathways <- Ingenuity_Canonical_Pathways[, c(1, 3, 4, 2)] %>% 
      subset(Ingenuity_Canonical_Pathways$Neg.log.10.p >= 1.3)
    
    Ingenuity_Canonical_Pathways <- Ingenuity_Canonical_Pathways %>% arrange(desc(Neg.log.10.p))
    
    # Set names of columns
    colnames(Ingenuity_Canonical_Pathways) <- c('Pathway',
                                                paste0(as.character(Individual_file), ' zScore'),
                                                paste0(as.character(Individual_file), ' Ratio'),
                                                paste0(as.character(Individual_file), ' Neg.log.10.p'))
    
    # Save the results to Canonical list
    Ingenuity_Canonical_Pathways <- Ingenuity_Canonical_Pathways[1:10, ]
    
  })
  
  names(Top_pathways) <- names(IPA_Results[[IPA_Grouping]])
  
  Top_pathways
  
})

# Rename the list to match the groups
names(Top_pathways) <- Groupings
```

### Isolate the pathway z-scores for each of the top 10 pathways isolated above. NA z-scores represent pathways for which z-scores could not be calculated.

```{r}
zScore <- lapply(Groupings, function(IPA_Grouping){

  # Merge top pathways removing repeats
  merge <- Reduce(function(x, y) merge(x, y, all = TRUE), Top_pathways[[IPA_Grouping]])
  Pathways <- merge$Pathway
  
  # Merge all canonical pathway results
  zScore <- Reduce(function(x, y) merge(x, y, all = TRUE), Canonical[[IPA_Grouping]]) 
  
  # Isolate Z scores
  zScore <- data.frame(zScore[, seq(from = 2, 
                                    to = ncol(zScore) - 1, 
                                    by = 3)], 
                       row.names = zScore$Pathway) 
  
  # Save Z scores for top pathways
  zScore <- zScore[Pathways, ] 
  
  # Remove Z scores with all NA values
  Remove <- zScore %>% filter_all(all_vars(is.na(.))) %>% rownames()
  zScore <- zScore[!(row.names(zScore) %in% Remove), ]
  
  # Calculate variance between groups and remove pathways with no variance between samples
  var <- as.data.frame(sapply(as.data.frame(t(zScore)), var))
  var <- var %>% subset(var[, 1] == 0)
  zScore <- zScore[!(row.names(zScore) %in% row.names(var)), ]
  
  # Rename columns to match desired output
  if(IPA_Grouping == 'By Polarization State'){

    colnames(zScore) <- gsub(pattern = '\\.zScore',
                                  replacement = '',
                                  x = colnames(zScore))
    
    colnames(zScore) <- gsub(pattern = '\\.',
                             replacement = ' ',
                             x = colnames(zScore))
  }

  else{
    if(IPA_Grouping == 'M1 vs M0' |
       IPA_Grouping == 'M2 vs M0'){

      colnames(zScore) <- gsub(pattern = '\\.zScore',
                               replacement = '',
                               x = colnames(zScore))
      
      colnames(zScore) <- gsub(pattern = 'M\\.',
                               replacement = 'M(',
                               x = colnames(zScore))
      
      colnames(zScore) <- gsub(pattern = '\\.',
                               replacement = ')',
                               x = colnames(zScore))

    }

    else{
      
      colnames(zScore) <- gsub(pattern = '\\.zScore',
                               replacement = '',
                               x = colnames(zScore))
      
      colnames(zScore) <- gsub(pattern = 'M\\.',
                               replacement = 'M(',
                               x = colnames(zScore))
      
      colnames(zScore) <- gsub(pattern = 'vs\\.',
                               replacement = 'vs ',
                               x = colnames(zScore))
      
      colnames(zScore) <- gsub(pattern = '\\.',
                               replacement = ')',
                               x = colnames(zScore))
      
      colnames(zScore) <- gsub(pattern = '))',
                               replacement = ') ',
                               x = colnames(zScore))
      
      }
  }

  zScore
  
})

names(zScore) <- Groupings

save(IPA_Results, 
     Top_pathways, 
     file = 'IPA_Results.RData')
```

### Generate heatmap of the resulting z-scores

```{r}
# Set color palette for heatmap
myCol <- colorRamp2(c(-5, 0, 5), hcl_palette = "Vik")

lapply(Groupings, function(IPA_Grouping){
  
  Heatmap_data <- as.matrix(zScore[[IPA_Grouping]])

  hmap <- Heatmap(Heatmap_data,
                   
                   cluster_row_slices = FALSE,
                   gap = unit(0.75, "mm"),
                   border = TRUE,
                   width = ncol(Heatmap_data) * unit(50, 'mm'), 
                   height = ncol(Heatmap_data) * unit(125, 'mm'),
                   
                   name = 'Canonical Pathway\nZ-Score',
                   
                   # Color scheme
                   col = myCol,
                   
                   # Set color for NA values
                   na_col = "pink",
                   
                   # Add values to boxes
                   # Use black or white text based on value
                   cell_fun = function(j, i, x, y, width, height, fill) {
                     grid.text(sprintf("%.1f", 
                                       Heatmap_data[i, j]), x, y, 
                               gp = gpar(fontsize = 24, 
                                         col = if_else(Heatmap_data[i, j] < 3 & 
                                                         Heatmap_data[i, j] > -3 | 
                                                         is.na(Heatmap_data[i, j]) == TRUE, 
                                                       'black', 
                                                       'white'),
                                         fontface = 'bold'))},
                   
                   # parameters for the color-bar that represents gradient of expression
                   heatmap_legend_param = list(
                     color_bar = 'continuous',
                     at = c(-5, -2.5, 0, 2.5, 5),
                     legend_direction = 'horizontal',
                     legend_width = unit(15, 'cm'),
                     legend_height = unit(10.0, 'cm'),
                     title_position = 'topcenter',
                     title_gp = gpar(fontsize = 24, fontface = 'bold'),
                     labels_gp = gpar(fontsize = 20, fontface = 'bold')),
                   
                   # row (gene) parameters
                   cluster_rows = FALSE,
                   show_row_dend = FALSE,
                   row_title_side = 'right',
                   row_title_gp = gpar(fontsize = 12,  fontface = 'bold'),
                   show_row_names = TRUE,
                   row_names_gp = gpar(fontsize = 20, fontface = 'bold'),
                   row_names_side = 'right',
                   row_dend_width = unit(25,'mm'),
                   
                   # column (sample) parameters
                   cluster_columns = FALSE,
                   show_column_dend = FALSE,
                   column_title = '',
                   column_title_side = 'bottom',
                   column_title_gp = gpar(fontsize = 12, fontface = 'bold'),
                   show_column_names = TRUE,
                   column_names_gp = gpar(fontsize = 30, fontface = 'bold'),
                   column_names_rot = 45,
                   column_names_max_height = unit(10, 'cm'),
                   column_dend_height = unit(25,'mm'))
  
  tiff(file = paste0(IPA_Grouping, " Heatmap.tiff"), 
       height = ifelse(IPA_Grouping == 'M2 vs M2', 
                       2500,
                       2000),
       width = 2500)
  
  draw(hmap,
       heatmap_legend_side = 'top')
  
  dev.off()

})
```