---
title: "Differential Expression by Polarization State"
author: "Timothy Smyth"
date: "2024-03-27"
output: html_document
---

# Differential Expression by Polarization State

### This file performs differential expression analysis of monocyte-derived macrophage count data by polarization state using limma-voom. Raw count data of gene and sample filtered MDM data from 'Gene and Sample Filtering' is the input data for this analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Environment setup

```{r message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

# Load packages
library(tidyverse) # for data cleaning
library(dplyr) 
library(broom)
library(tibble)
library(edgeR)
library(limma)
library(EnhancedVolcano)
library(eulerr)
```

### Load and format count data

```{r}
# This is gene and sample filtered but not limma-voom corrected data
# The model written below takes batch (series) into account
load("Selected_Count_Data.RData")

# Isolate sample info and set to factor
info$polarization <- factor(info$polarization, 
                            levels = c("M0", "M1", "M2"))

info$series <- factor(info$series_id)

# Define factor objects for models
series_id <- factor(info$series)
polarization <- factor(info$polarization, levels = c("M0", "M1", "M2"))

# Create DGEList object
dataset <- DGEList(t(as.matrix(MDM_df)), 
                   group = polarization)

dataset <- calcNormFactors(dataset)

# Print the dimensions of the new object (Genes x Samples)
dim(dataset)
```

### Create model matrix, run voom, and fit linear models

```{r}
# Create a model with polarization and series as factors
model <- model.matrix(~0 + polarization + series_id)

colnames(model)[1:3] <- c("M0", "M1", "M2")

# Check if any coefficients in model cannot be estimated
# Return value of "NULL" means model has full column rank
nonEstimable(model.matrix(~0 + polarization + series_id))

# Run voom and plot mean-variance trend
# voom function automatically converts counts to logCPM
# by extracting library size and calculating normalization factors
y <- voom(dataset, 
          design = model, 
          plot = T)

# Fit linear model using weighted least squares for genes
fit <- lmFit(y, 
             model)
```

### Calculate differentially expressed genes between groups as defined by makeContrasts using contrast.fit, perform empirical Bayes smoothing to gene-wise standard errors with eBayes, and present statistics using topTable

```{r}
# Make a list of contrasts
list <- list(makeContrasts(M1-M0, 
                           levels = model),
             
             makeContrasts(M2-M0, 
                           levels = model),
             
             makeContrasts(M1-M2, 
                           levels = model))

# Calculate DEGs
DEGs <- lapply(list, function(x){
  
  tmp <- contrasts.fit(fit, x)
  tmp <- eBayes(tmp)
  
  topTable(tmp, sort.by = "P", n = Inf)
  
})

# Results are the following information:

# logFC: log2 fold change of group1/group2
# AveExpr: Average expression across all samples, in log2 CPM
# t: logFC divided by its standard error
# P.Value: Raw p-value (based on t) from test that logFC differs from 0
# adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value
# B: log-odds that gene is DE

# Isolate DEGs
list <- as.list(1:length(DEGs))

list <- lapply(list, function(x){
  
  tmp <- data.frame(cbind(DEGs[[x]][["logFC"]],
                          DEGs[[x]][["P.Value"]],
                          DEGs[[x]][["adj.P.Val"]],
                          DEGs[[x]]["AveExpr"]),
                    row.names = rownames(DEGs[[x]])) 
  
  colnames(tmp) <- c('log2FC', 'P.Value', 'adj.P.Val', 'Expression')
  
  tmp
  
})

names(list) <- c('M1_vs_M0', 'M2_vs_M0', 'M1_vs_M2')
```

### Create volcano plots describing differential expression results with BH adjusted p-values of 0.05 and log2 fold change values >= 2 or <= -2 as significance cutoffs. Add text and arrows describing the number of up or downregulated genes.

```{r}
volcano <- list()

# Make these in a for loop for easy graph naming
for(i in 1:length(list)){
  
  # Determine the number of significantly upregulated genes
  up <- list[[i]] %>% count(log2FC > 2 &
                              adj.P.Val < 0.05)
  
  up <- as.numeric(up[2, 2])
  
  # Determine the number of significantly downregulated genes
  down <- list[[i]] %>% count(log2FC < -2 &
                                adj.P.Val < 0.05)
  
  down <- as.numeric(down[2, 2])
  
  volcano[[i]] <- EnhancedVolcano(list[[i]],
                                  lab = NA,
                                  x = 'log2FC',
                                  y = 'adj.P.Val',
                                  FCcutoff = 2,
                                  pCutoff = 0.05,
                                  title = names(list)[[i]],
                                  pointSize = 1) 
  
  # Determine the y axis limit
  max <- max(volcano[[i]][["plot_env"]][["ylim"]])
  
  # Add arrows and numbers describing the number of up or downregulated genes
  volcano[[i]] <- volcano[[i]] + 
    
    # Set x axis from -10 to 10
    ggplot2::coord_cartesian(xlim = c(-10, 10)) +
    
    # Place downregulated numbers on plot
    ggplot2::geom_text(x = -6.5,
                       y = max* 0.95,
                       label = paste0(down),
                       size = 8.5) +
    
    # Place downregulation arrow on plot
    ggplot2::geom_segment(x = -7.8,
                          y = max,
                          xend = -7.8,
                          yend = max * 0.9,
                          arrow = arrow(length = unit(2, "mm")), 
                          linewidth = 1) +
    
    # Place upregulated numbers on plot
    ggplot2::geom_text(x = 7.5,
                       y = max * 0.95,
                       label = paste0(up),
                       size = 8.5) + 
    
    # Place upregulation arrow on plot
    ggplot2::geom_segment(x = 6.2,
                          y = max * 0.9,
                          xend = 6.2,
                          yend = max,
                          arrow = arrow(length = unit(2, "mm")), 
                          linewidth = 1)
  
  # Save the resulting plot
  tiff(paste0(names(list)[[i]], '.tiff'),
       units = "cm",
       width = 18,
       height = 16,
       res = 300)
  
  print(volcano[[i]])
  
  dev.off()
  
}
```

### Save the DEG results

```{r}
# Isolate genes with adj.P.Val < 0.05 for IPA
DEGs <- lapply(list, function(x){
  
  tmp <- x %>% subset(adj.P.Val < 0.05 & (log2FC >= 2 | log2FC <= -2))
  
  # Change . back to - for gene names as IPA expects -
  row.names(tmp) <- gsub('\\.', "\\-", row.names(tmp))
  
  tmp
  
})

names(DEGs) <- c('M1', 'M2', 'M1_M2')

# Export DEGs [Log2FC and BH adjusted p values]
write.csv(DEGs[['M1']], file = 'M1.csv')
write.csv(DEGs[['M2']], file = 'M2.csv')
write.csv(DEGs[['M1_M2']], file = 'M1_M2.csv')
```

### Create Euler plots describing overlap of differentially expressed genes between groups

```{r}
Diff <- lapply(DEGs, function(x){
  
  up <- x %>% subset(log2FC >= 2) %>% rownames_to_column()
  down <- x %>% subset(log2FC <= 2) %>% rownames_to_column()
  
  x <- list(up, down)
  names(x) <- c('Up', 'Down')
  x
  
})

names(Diff) <- c('M1', 'M2', 'M1_M2')

Direction <- list('Up', 'Down')

lapply(Direction, function(y){
  
  eul_list <- list(unique(Diff[[1]][[y]]$rowname),
                   unique(Diff[[2]][[y]]$rowname),
                   unique(Diff[[3]][[y]]$rowname))
  
  names(eul_list) <- c(names(Diff))
  
  col <- c('red', 
           'yellow', 
           'blue')
  
  fit <- eulerr::euler(eul_list)
  
  fit <- plot(fit,
              fills = list(fill = col, alpha = 0.5),
              legend = list(side = 'bottom', col = col, text.col = 'black'),
              quantities = list(cex = 2), 
              main = paste0(y, 'regulated ', 'Genes'), 
              xlim = c(-15, 15), 
              ylim = c(-15, 15))
  
  tiff(paste0(y, 'regulated ', 'Genes.tiff'),
       units = "in",
       width = 7,
       height = 5,
       res = 300)
  print(fit)
  dev.off()
})
```

