---
title: "Differential Expression Comparisons and Euler Plots"
author: "Timothy Smyth"
date: "2024-03-27"
output: html_document
---

# Differential Expression Comparisons and Euler Plots

### This file creates Euler plots for each comparison of interest (M1 vs M0, M1 vs M1, etc) for Figures 1-3 and Supplemental Figures S2-4

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Environment setup

```{r, message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

# Load packages
library(tidyverse) # for data cleaning
library(dplyr) 
library(readxl)
library(ComplexHeatmap)
library(circlize)
library(openxlsx)
library(eulerr)
```

### Load data

```{r}
load("Method_DEG_Results.RData")
load("Pol_State_DEG_Results.RData")
load("IPA_Results.RData")

DEG_results <- c(Pol_DEG_Results, Method_DEG_Results)

Groupings <- list('By Polarization State', 
                  'M1 vs M0', 
                  'M1 vs M1', 
                  'M2 vs M0', 
                  'M2 vs M2')

names(DEG_results) <- gsub(pattern = ') vs M0', 
                           replacement = ')', 
                           x = names(DEG_results))
```

### Determine overlapping genes between each subset in each comparison (i.e. M(LPS) vs M0 for Figure 2, M(LPS) vs M(IFNg) for Supplemental Figure S4) for IPA

```{r}
IPA_Genes <- lapply(Groupings, function(IPA_Grouping){
  
  # Save the names of each individual group within larger groupings
  names <- names(Top_pathways[[IPA_Grouping]])
  
  ########## Chunk 1: Isolate top 10 IPA canonical pathways and their input genes for each group ##########
  
  IPA_pathways <- lapply(Top_pathways[[IPA_Grouping]], function(Individual_group){
    
    Individual_group$Pathway
    
  })
  
  IPA_pathways <- unique(as.character(do.call(rbind, IPA_pathways))) %>% na.omit()

  Top_pathway_molecules <- lapply(names(IPA_Results[[IPA_Grouping]]), function(Individual_group){
    
    Top_pathway_molecules <- data.frame(IPA_Results[[IPA_Grouping]][[Individual_group]][, c('Ingenuity Canonical Pathways', 'Molecules')],
                                        row.names = IPA_Results[[IPA_Grouping]][[Individual_group]][, 'Ingenuity Canonical Pathways'])
    
    Top_pathway_molecules <- Top_pathway_molecules[IPA_pathways, ]
    Top_pathway_molecules[is.na(Top_pathway_molecules)] <- 'NOT SIGNIFICANT'
    rownames(Top_pathway_molecules) <- IPA_pathways
    Top_pathway_molecules
    
  })
  
  names(Top_pathway_molecules) <- names
  
  ########## Chunk 2: Isolate significantly DE genes in each significant IPA pathway ##########

  DE_Genes <- lapply(names, function(Group_name){

    Genes <- DEG_results[[Group_name]] %>% subset(abs(log2FC) >= 2 & adj.P.Val < 0.05)
    Genes <- Genes[, 'HUGO_IDs']
    
    Save_Genes <- list()

    for(i in 1:nrow(Top_pathway_molecules[[Group_name]])){
      
      if(Top_pathway_molecules[[Group_name]][i, 'Molecules'] == 'NOT SIGNIFICANT'){
        
        Save_Genes[[i]] <- 'NOT SIGNIFICANT'
        
      }
      
      else{
        
        Pathway_Genes <- unlist(strsplit(Top_pathway_molecules[[Group_name]][i, 'Molecules'], ","))

        keep = Genes %in% Pathway_Genes
        keep = which(keep == TRUE)
      
        ifelse(length(keep) > 0,
               Save_Genes[[i]] <- list(unique(Genes[keep])),
               Save_Genes[[i]] <- 'NO GENES')
      }

    }

    # Remove unwanted characters
    Save_Genes <- as.character(do.call(rbind, Save_Genes)) %>% 
      str_remove_all("[^[:UPPER:][[:digit:]], ]") %>% 
      gsub(pattern = '\\,', replace = '\\, ') 
    
    DE_Genes <- data.frame(Save_Genes, 
                           row.names = rownames(Top_pathway_molecules[[Group_name]]))
    
    colnames(DE_Genes) <- Group_name
    
    DE_Genes
    
  })
  
  names(DE_Genes) <- names
  
  DE_Genes <- data.frame(DE_Genes[1:length(DE_Genes)]) %>% 
    rownames_to_column() %>% 
    setNames(c('Pathway', names))
  
  # Rename columns to match desired output
    if(IPA_Grouping == 'By Polarization State'){
    
    colnames(DE_Genes) <- gsub(pattern = '\\.',
                             replacement = ' ',
                             x = colnames(DE_Genes))
    }
  
  if(IPA_Grouping == 'M1 vs M0' |
     IPA_Grouping == 'M2 vs M0'){
    
    colnames(DE_Genes) <- gsub(pattern = 'M\\.',
                             replacement = 'M(',
                             x = colnames(DE_Genes))
    
    colnames(DE_Genes) <- gsub(pattern = '\\.',
                             replacement = ')',
                             x = colnames(DE_Genes))

  }

  else{
    
    colnames(DE_Genes) <- gsub(pattern = 'M\\.',
                             replacement = 'M(',
                             x = colnames(DE_Genes))
    
    colnames(DE_Genes) <- gsub(pattern = 'vs\\.',
                             replacement = 'vs ',
                             x = colnames(DE_Genes))
    
    colnames(DE_Genes) <- gsub(pattern = '\\.',
                             replacement = ')',
                             x = colnames(DE_Genes))
    
    colnames(DE_Genes) <- gsub(pattern = '))',
                             replacement = ') ',
                             x = colnames(DE_Genes))
    
  }
  
  DE_Genes

})

names(IPA_Genes) <- Groupings
```

### Isolate differentially expressed genes and sort based on upregulated or downregulated

```{r}
# Isolate DEGs
All_DEGs <- lapply(DEG_results, function(Dataset){
  
  # Isolate upregulated genes and rename Ensembl column
  up <- Dataset %>% subset(log2FC >= 2 & adj.P.Val < 0.05) %>% rownames_to_column()
  colnames(up)[1] <- 'Ensembl_ID'
  up <- up[, c(1, 6, 2:5)]

  # Isolate downregulated genes and rename Ensembl column
  down <- Dataset %>% subset(log2FC <= -2 & adj.P.Val < 0.05) %>% rownames_to_column()
  colnames(down)[1] <- 'Ensembl_ID'
  down <- down[, c(1, 6, 2:5)]
  
  # Save up and downregulated genes in a list and rename
  Dataset <- list(up, down)
  names(Dataset) <- c('Upregulated', 'Downregulated')
  Dataset
  
})
```

### Isolate gene names for significantly differentially expressed inflammasome pathway linked genes

```{r}
Inflammasome <- read.csv('Inflammasome_List.csv')

# As some HUGO names map to multiple Ensembl IDs, determine all Ensembl IDs in pathway and 
# match up Ensembl IDs with HUGO IDs
Inflammasome <- data.frame('Gene_ID' = rownames(Symbol)[which(Symbol$Genes.gene_name %in% 
                                                                Inflammasome$HUGO_ID == TRUE)], 
                           
                           'HUGO_ID' = Symbol[which(Symbol$Genes.gene_name %in% 
                                                                Inflammasome$HUGO_ID == TRUE), 
                                              'Genes.gene_name'])

Inflammasome <- lapply(names(All_DEGs), function(Dataset_name){
  
  Inflammasome <- lapply(c('Upregulated', 'Downregulated'), function(Dataset){
    
    keep <- which(All_DEGs[[Dataset_name]][[Dataset]]$Ensembl_ID %in% Inflammasome$Gene_ID == TRUE)
    
    genes <- data.frame('Ensembl_ID' = All_DEGs[[Dataset_name]][[Dataset]]$Ensembl, 
                        'HUGO_IDs' = All_DEGs[[Dataset_name]][[Dataset]]$HUGO_IDs, 
                        'log2FC' = All_DEGs[[Dataset_name]][[Dataset]]$log2FC,
                        'adj.P.Val' = All_DEGs[[Dataset_name]][[Dataset]]$adj.P.Val)
    
    genes <- data.frame(genes[keep, ])
    
    genes
    
  })
  
  names(Inflammasome) <- c('Upregulated', 'Downregulated')
  Inflammasome
  
})

names(Inflammasome) <- names(All_DEGs)
```

### Isolate gene names for significantly differentially expressed transcription factors

```{r}
# https://humantfs.ccbr.utoronto.ca/download.php
Transcription_Factors <- read.csv('Transcription_Factor_List.csv')

# As some HUGO names map to multiple Ensembl IDs, determine all Ensembl IDs in pathway and 
# match up Ensembl IDs with HUGO IDs
Transcription_Factors <- data.frame('Gene_ID' = rownames(Symbol)[which(rownames(Symbol) %in% 
                                                                         Transcription_Factors$Gene_ID == TRUE)], 
                                    
                                    'HUGO_ID' = Symbol[which(rownames(Symbol) %in% 
                                                               Transcription_Factors$Gene_ID == TRUE), ])

TFs <- lapply(names(All_DEGs), function(Dataset_name){
  
  TFs <- lapply(c('Upregulated', 'Downregulated'), function(Dataset){
    
    keep <- which(All_DEGs[[Dataset_name]][[Dataset]]$Ensembl_ID %in% Transcription_Factors$Gene_ID == TRUE)
    
    genes <- data.frame('Ensembl_ID' = All_DEGs[[Dataset_name]][[Dataset]]$Ensembl, 
                        'HUGO_IDs' = All_DEGs[[Dataset_name]][[Dataset]]$HUGO_IDs, 
                        'log2FC' = All_DEGs[[Dataset_name]][[Dataset]]$log2FC,
                        'adj.P.Val' = All_DEGs[[Dataset_name]][[Dataset]]$adj.P.Val)
    
    genes <- data.frame(genes[keep, ])
    
    genes
    
  })
  
  names(TFs) <- c('Upregulated', 'Downregulated')
  TFs
  
})

names(TFs) <- names(All_DEGs)
```

### Reformat above data

```{r}
# Function to reformat data for export
Format_dataset <- function(Dataset, Data_grouping){
  
  Formatted_data <- lapply(Data_grouping, function(x){
    
    names <- names(Top_pathways[[x]])
    
    Formatted_data <- lapply(names, function(y){

      Dataset[[y]]

    })
    
    names(Formatted_data) <- names
    
    Formatted_data
    
  })
  
}

All_DEGs <- Format_dataset(Dataset = All_DEGs, Data_grouping = Groupings)
names(All_DEGs) <- Groupings

Inflammasome <- Format_dataset(Dataset = Inflammasome, Data_grouping = Groupings)
names(Inflammasome) <- Groupings

TFs <- Format_dataset(Dataset = TFs, Data_grouping = Groupings)
names(TFs) <- Groupings

All_data <- list(All_DEGs, 
                 Inflammasome, 
                 TFs)

names(All_data) <- c('All_DEGs', 
                     'Inflammasome', 
                     'TFs')
```

### Create Euler plots describing overlap of differentially expressed genes between groups

```{r}
Gene_overlap <- lapply(names(All_data), function(Primary_dataset){
  
  Primary_dataset_name <- Primary_dataset

  # Track if overall DEG dataset or other dataset is input
  # This will control whether euler plots are created
  ifelse(Primary_dataset_name == 'All_DEGs',
         Is_All_DEGs <- 'Yes',
         Is_All_DEGs <- 'No')
  
  ####################### Chunk 1: Match up/downregulated genes in comparisons #######################
   
  # Rearrange data so up/downregulated data for each subgroup (pol state/method) within
  # same comparison is within the same sublist for further processing
  Match_DEG_Directionality <- lapply(All_data[[Primary_dataset]], function(Pol_state_or_method){
    
    Pol_state_or_method_names <- names(Pol_state_or_method)

    # Create named lists to store data
    Upregulated_Ensembl <- vector("list", length(Pol_state_or_method))
    names(Upregulated_Ensembl) <- Pol_state_or_method_names

    Upregulated_HUGO <- vector("list", length(Pol_state_or_method))
    names(Upregulated_HUGO) <- Pol_state_or_method_names

    Downregulated_Ensembl <- vector("list", length(Pol_state_or_method))
    names(Downregulated_Ensembl) <- Pol_state_or_method_names

    Downregulated_HUGO <- vector("list", length(Pol_state_or_method))
    names(Downregulated_HUGO) <- Pol_state_or_method_names
    
    # For each subgroup (pol state/method), save the up/downregulated data
    # in their respective storage list
    for(i in 1:length(Pol_state_or_method)){

      Upregulated_Ensembl[[i]] <- Pol_state_or_method[[i]][['Upregulated']]$Ensembl_ID
      Upregulated_HUGO[[i]] <- Pol_state_or_method[[i]][['Upregulated']]$HUGO_ID

      Downregulated_Ensembl[[i]] <- Pol_state_or_method[[i]][['Downregulated']]$Ensembl_ID
      Downregulated_HUGO[[i]] <- Pol_state_or_method[[i]][['Downregulated']]$HUGO_ID

    }
    
    # Reorganize up/downregulated DEG data into new list and rename to match
    Match_DEG_Directionality <- list(Upregulated_Ensembl,
                                     Upregulated_HUGO,
                                     Downregulated_Ensembl,
                                     Downregulated_HUGO)

    # Rename elements to match stored data information
    names(Match_DEG_Directionality) <- c('Upregulated_Ensembl',
                                         'Upregulated_HUGO',
                                         'Downregulated_Ensembl',
                                         'Downregulated_HUGO')

    Match_DEG_Directionality
    
  })
  
  ####################### Chunk 2: Determine overlapping DEGs between pol groups/methods #######################

  Overlapping_DEGs <- lapply(Groupings, function(Group){
    
    Data <- Match_DEG_Directionality[[Group]]

    Overlapping_DEGs <- lapply(c('Up', 'Down'), function(Direction){
      
      Up_or_down <- paste0(Direction, 'regulated_Ensembl')
      Plot_name <- paste0(Direction, 'regulated Genes')
      
      Overlapping_DEGs <- lapply(c('regulated_Ensembl', 'regulated_HUGO'), function(x){
        
        Gene_type <- paste0(Direction, x)

        ####################### Chunk 3: Create Euler plot determining overlap of all DEGs #######################
        
        if(Is_All_DEGs == 'Yes'){
  
          # Define colors for euler plots
          colors <- c('blue3',
                      'khaki3',
                      'firebrick3',
                      'purple3',
                      'seagreen3',
                      'darkorange3')
    
          # Set group colors
          col <- colors[1:length(colors)]
    
          # Fit the Euler plot using Eulerr package
          fit <- eulerr::euler(Data[[Up_or_down]])
          
          # Create Euler plot
          fit <- plot(fit,
                      fills = list(fill = col, alpha = 0.5),
                      legend = list(side = 'bottom', col = col, text.col = 'black'),
                      quantities = list(cex = 1.5),
                      main = paste0(Plot_name),
                      xlim = c(-15, 15),
                      ylim = c(-15, 15))
    
          # Export as a tiff
          tiff(paste0(Direction, 'regulated ', Group, ' .tiff'),
               units = "in",
               width = 7,
               height = 5.5,
               res = 300)
          print(fit)
          dev.off()
          
        }
        
        ####################### Chunk 4: Create data frame containing overlapping information #######################
        
        # Defining overlap of genes can be easily done using ggVennDiagram functions
        Overlap_venn <- ggVennDiagram:::Venn(Data[[Gene_type]]) %>% ggVennDiagram::process_data()

        # Isolate genes unique to each group and combine to one data frame, removing NA values
        # generated from combining unequal length vectors into one data frame
        Gene_overlap <- Overlap_venn@region[["item"]]
        Gene_overlap <- do.call(qpcR:::cbind.na, Gene_overlap) %>% data.frame()
        Gene_overlap[is.na(Gene_overlap)] <- ''

        # Rename to define overlapping groups
        names(Gene_overlap) <- Overlap_venn@region[["name"]] %>%
          gsub(pattern = '..',
               replacement = '_Overlap_',
               fixed = TRUE)
        
        Gene_overlap
        
      })
      
      names(Overlapping_DEGs) <- c('Ensembl',
                                   'HUGO')
      
      Overlapping_DEGs
      
    })
    
    names(Overlapping_DEGs) <- c('Upregulated',
                                 'Downregulated')

    Overlapping_DEGs
    
  })
  
  names(Overlapping_DEGs) <- Groupings
  
  Overlapping_DEGs
  
})

names(Gene_overlap) <- names(All_data)
```

### Export DE and gene overlap data for Supplemental Files 1-6

```{r}
Export_data <- lapply(Groupings, function(Individual_Grouping){
  
  # Isolate top IPA pathways and DE genes which were included by each group in the IPA analysis
  Genes_in_IPA_Analysis <- list(IPA_Genes[[Individual_Grouping]])
  
  # Reorganize all data for excel file writing
  Reorganize_data <- lapply(names(All_data), function(Primary_dataset){
    
    # All DEGs dataset has DE results, so export the raw DEG results with it
    if(Primary_dataset == 'All_DEGs'){
      
      Raw_DEG_dataset <- lapply(names(Top_pathways[[Individual_Grouping]]), function(DEG_Dataset){
        
        Raw_DEG_dataset <- DEG_results[[DEG_Dataset]] %>% rownames_to_column()
        colnames(Raw_DEG_dataset)[c(1, 6)] <- c('Ensembl_ID', 'HUGO_ID')
        Raw_DEG_dataset

      })
      
      # Match names
      names(Raw_DEG_dataset) <- names(Top_pathways[[Individual_Grouping]])
      
      # Isolate upregulated and downregulated overlapping DE data
      Overlap_Data_Upregulated <- Gene_overlap[[Primary_dataset]][[Individual_Grouping]][['Upregulated']]
      Overlap_Data_Downregulated <- Gene_overlap[[Primary_dataset]][[Individual_Grouping]][['Downregulated']]

      # Compile data into one place and set names to match data
      Reorganize_data <- c(Raw_DEG_dataset, 
                           Genes_in_IPA_Analysis,
                           Overlap_Data_Upregulated, 
                           Overlap_Data_Downregulated) %>% setNames(c(names(Raw_DEG_dataset), 
                                                                      'IPA_Analyzed_Genes',
                                                                      paste0(Primary_dataset, '_Upreg_Ensembl'), 
                                                                      paste0(Primary_dataset, '_Upreg_HUGO'), 
                                                                      paste0(Primary_dataset, '_Downreg_Ensembl'), 
                                                                      paste0(Primary_dataset, '_Downreg_HUGO')))
      
    }
    
    else{
      
      # Isolate upregulated and downregulated overlapping DE data
      Overlap_Data_Upregulated <- Gene_overlap[[Primary_dataset]][[Individual_Grouping]][['Upregulated']]
      Overlap_Data_Downregulated <- Gene_overlap[[Primary_dataset]][[Individual_Grouping]][['Downregulated']]
      
      # Compile data into one place and set names to match data
      Reorganize_data <- c(Overlap_Data_Upregulated, 
                           Overlap_Data_Downregulated) %>% setNames(c(paste0(Primary_dataset, '_Upreg_Ensembl'), 
                                                                      paste0(Primary_dataset, '_Upreg_HUGO'), 
                                                                      paste0(Primary_dataset, '_Downreg_Ensembl'),
                                                                      paste0(Primary_dataset, '_Downreg_HUGO')))
      
    }
    
  })

  # Combine all reorganized data into one place
  Reorganize_data <- c(Reorganize_data[[1]], 
                       Reorganize_data[[2]], 
                       Reorganize_data[[3]])
  
  # Create an excel workbook
  Excel_workbooks <- createWorkbook()

  # Write each group as a separate worksheet and save as an excel file
  Excel_workbooks <- lapply(names(Reorganize_data), function(Dataset){

    # Add data to a new worksheet
    addWorksheet(Excel_workbooks,
                 Dataset)

    # Write new worksheet to created excel file
    writeDataTable(Excel_workbooks,
                   Dataset,
                   x = Reorganize_data[[Dataset]],
                   rowNames = F)

    # Save the created workbook as a single excel file
    saveWorkbook(Excel_workbooks,
                 paste0(Individual_Grouping,
                        ' DEGs.xlsx'),
                 overwrite = TRUE)

  })
})
```