---
title: "Top 1000 Genes Analysis"
author: "Timothy Smyth"
date: "2024-03-27"
output: html_document
---

# Top 1000 Genes Analysis

This file generates a principal component analysis plot and heatmap visualizing gene expression patterns of M0 and M1 macrophages using the top 1000 genes as determined by random forest permutation importance.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Environment setup

```{r message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

library(tibble)
library(tidyverse)
library(dplyr) 
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
library(digest)
library(cluster)
library(factoextra)
library(parallelpam)
library(openxlsx)
```

### Load data and generate PCA plot

```{r}
# Load data
load('Top_1000_RF_Genes.RData')

# Save metadata
info <- MDM_df[1:3]

# Retain only top 1000 genes by variance
MDM_df <- MDM_df[, top_1000]

# Ensure importance results correctly formatted
Importance_results <- Importance_results[top_1000, ]

######### PCA ######### 

# Run PCA
pca.res <- prcomp(MDM_df, center = TRUE, scale = TRUE)

# Set theme
theme_set(theme_bw())

# Set colors
col2 <- c('blue',          
          'gray70',
          'skyblue2',
          'maroon2')

info$polarization <- factor(info$polarization, 
                            levels = c('M0', 'M1', 'M2'))

info$ID <- factor(info$ID,
                    levels = c('M0', 
                               'IFN', 'LPS', 'LPS_IFN',
                               'IL4', 'IL10', 'IL13', 'IL4_IL13'))

# Set value for shapes to be used below
shapes <- c(18, 9, 15, 12, 17, 2, 19, 13, 
            
            1, 3, 4, 5, 6, 7, 8, 10, 11, 
            
            12, 13, 14, 16, 18, 20, 
            
            18, 9, 15, 12, 17, 2, 19, 13)

# All groups cluster plot
fviz_pca_ind(pca.res,
             label = "none",
             habillage = info$ID,
             palette = col2,
             addEllipses = TRUE,
             ellipse.alpha = 0.15,
             point.size = 0.5) + 
  
  ggtitle("PCA - Clustering") +
  ylim(-20, 20) +
  
  theme(axis.title = element_text(size = 14), 
        axis.text = element_text(size = 12), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        panel.border = element_rect(fill = NA, color = "black", size = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) 
```

### Perform partition around medoid (PAM) clustering for samples and genes

```{r}
# Partitioning around medoids using parallelpam for samples
parallelpam::JWriteBin(as.matrix(t(scale(t(MDM_df)))),
                       "MDM_results_group.bin",
                       dtype = "float",
                       dmtype = "full")

CalcAndWriteDissimilarityMatrix("MDM_results_group.bin", 
                                "MDM_results_group2.bin",
                                distype = "L2", 
                                restype = "float",
                                comment = "L2 distance for vectors in jmatrix file vst_results_group.bin", 
                                nthreads = 8)

JMatInfo("MDM_results_group2.bin")

MDM_results_group_pam = ApplyPAM("MDM_results_group2.bin", 
                                 k = 4, # by polarization method
                                 init_method = "BUILD", 
                                 max_iter = 1000, 
                                 nthreads = 8)

#################################################################################

# Partitioning around medoids using parallelpam for genes
parallelpam::JWriteBin(as.matrix(t(scale(MDM_df))),
                       "MDM_results.bin",
                       dtype = "float",
                       dmtype = "full")

CalcAndWriteDissimilarityMatrix("MDM_results.bin", 
                                "MDM_results2.bin",
                                distype = "L2", 
                                restype = "float",
                                comment = "L2 distance for vectors in jmatrix file vst_results.bin", 
                                nthreads = 8)

JMatInfo("MDM_results2.bin")

MDM_results_pam = ApplyPAM("MDM_results2.bin", 
                           k = 4,
                           init_method = "BUILD", 
                           max_iter = 1000, 
                           nthreads = 8)
```

### Generate heatmap using results of PAM clustering

```{r}
# Scale results
heat <- scale(MDM_df)

pol_level <- paste0('Cluster ', seq_along(1:max(MDM_results_group_pam$clasif)))

levels <- paste0('Cluster ', seq_along(1:max(MDM_results_pam$clasif)))

# Define clusters calculated above for sorting
MDM_results_group_pam$clasif <- paste0('Cluster ', MDM_results_group_pam$clasif)
MDM_results_pam$clasif <- paste0('Cluster ', MDM_results_pam$clasif)

# Fix order of the clusters to have 1 to 4, top to bottom
MDM_results_group_pam$clasif <- factor(MDM_results_group_pam$clasif,
                                 levels = pol_level)

MDM_results_pam$clasif <- factor(MDM_results_pam$clasif,
                                 levels = levels)

# Set color palette
myCol <- colorRamp2(c(-2, 0, 2), hcl_palette = "Vik")

# Create annotation data frame
annotation <- data.frame(
  ID = info$ID, 
  Polarization = info$polarization,
  stringsAsFactors = FALSE)

colors <- list(
  ID = c('M0' = 'blue', 
         'LPS' = 'skyblue2',
         'IFN' = 'gray70',
         'LPS_IFN' = 'maroon2'),
  
  Polarization = c('M0' = 'blue', 
                   'M1' = 'red'))

# Define column annotation
colAnn <- HeatmapAnnotation(
  df = annotation,
  which = 'col', # 'col' (samples) or 'row' (gene) annotation
  col = colors,
  annotation_height = 0.6,
  annotation_width = unit(1, 'cm'),
  gap = unit(1, 'mm'),
  annotation_legend_param = list(
    Polarization = list(
      nrow = 2, # number of rows across which the legend will be arranged
      title = 'Polarization',
      title_position = 'topcenter',
      legend_direction = 'vertical',
      title_gp = gpar(fontsize = 10, fontface = 'bold'),
      labels_gp = gpar(fontsize = 10, fontface = 'bold'))))

heatmap <- Heatmap(t(heat),

                 # split the genes / rows according to the PAM clusters
                 row_split = MDM_results_pam$clasif,
                 column_split = MDM_results_group_pam$clasif,
                 cluster_row_slices = FALSE,
                 row_gap = unit(2.5, "mm"),
                 column_gap = unit(2.5, 'mm'),
                 border = TRUE,

                 name = 'Gene\nZ-Score\ ',

                 col = myCol,

                 # parameters for the color-bar that represents gradient of expression
                 heatmap_legend_param = list(
                   color_bar = 'continuous',
                   legend_direction = 'horizontal',
                   legend_width = unit(25, 'cm'),
                   legend_height = unit(25.0, 'cm'),
                   title_position = 'topcenter',
                   title_gp = gpar(fontsize = 24, fontface = 'bold'),
                   labels_gp = gpar(fontsize = 24, fontface = 'bold')),

                 # row (gene) parameters
                 cluster_rows = TRUE,
                 show_row_dend = TRUE,
                 row_title_side = 'left',
                 row_title_gp = gpar(fontsize = 24,  fontface = 'bold'),
                 row_title_rot = 0,
                 show_row_names = FALSE,
                 row_dend_width = unit(25,'mm'),

                 # column (sample) parameters
                 cluster_columns = TRUE,
                 show_column_dend = TRUE,
                 column_title = '',
                 column_title_side = 'bottom',
                 column_title_gp = gpar(fontsize = 24, fontface = 'bold'),
                 column_title_rot = 45,
                 show_column_names = FALSE,
                 column_dend_height = unit(25,'mm'),

                 # cluster methods for rows and columns
                 clustering_distance_columns = function(x) as.dist(1 - cor(t(x))),
                 clustering_method_columns = 'ward.D2',
                 clustering_distance_rows = function(x) as.dist(1 - cor(t(x))),
                 clustering_method_rows = 'ward.D2',

                 # specify top and bottom annotations
                 top_annotation = colAnn)

png(file = "Heatmap.png", height = 1250, width = 2000)

start.time <- Sys.time()

draw(heatmap,
     heatmap_legend_side = 'bottom',
     annotation_legend_side = 'top')

end.time <- Sys.time()
a <- end.time - start.time
a

dev.off()
```

### Export data regarding top 1000 genes by permutation importance

```{r}
# Calculate mean +/- SD of log count per million data
Mean_logCPM <- MDM_df %>% 
  group_by(info$ID) %>% 
  summarise_all(list(~ str_c(round(mean(.), 2), 
                             " ± ", 
                             round(sd(.), 2))))

# Rename sample ID
Mean_logCPM$`info$ID` <- paste0(Mean_logCPM$`info$ID`, 
                                ' Mean Log CPM')

# Set to data frame
Mean_logCPM <- data.frame(Mean_logCPM[-1], 
                          row.names = Mean_logCPM$`info$ID`)

# Transpose data frame and retain colnames
Mean_logCPM <- t(data.frame(as.matrix(Mean_logCPM)))

#########################

# Combine gene IDs, RF model importance values, gene PAM cluster info, and mean log cpm data
Importance_results <- data.frame('Ensembl_ID' = rownames(Importance_results), 
                                 Importance_results, 
                                 'Pam Cluster' = MDM_results_pam[["clasif"]],
                                 Mean_logCPM)

colnames(Importance_results) <- gsub(pattern = '\\.', 
                                     replacement = ' ', 
                                     x = colnames(Importance_results))

wb <- createWorkbook()

# Add data to a new worksheet
addWorksheet(wb,
             'Importance_results')

# Write new worksheet to created excel file
writeDataTable(wb,
               'Importance_results',
               x = Importance_results,
               rowNames = F)

# Save the created workbook as a single excel file
saveWorkbook(wb,
             'Importance Results.xlsx',
             overwrite = TRUE)
```